MS&E 226: Project Part 1 Code

### 2024-12-04 MS&E 226_Project Part 2_Code.Rmd
ols significant coeff

#### (a)
```{r 2a}
# Extract the final Ridge model using the best lambda
final_ridge_model <- ridge_model$finalModel

# Extract coefficients from the final model
coefficients <- coef(final_ridge_model, s = final_lambda_ridge)

# Print the coefficients
print(coefficients)

@@

#### (b)
```{r}
# Select a subset of covariates
subset_column_names <- c(  "Demand.Summer.Peak", "Demand.Winter.Peak", "Sources.Generation", "Utility.Region", "Max.Revenue.Sector", "Revenue.Total", "Utility.Type")

# Create a new data frame with the subset of columns
df_subset <- df[, subset_column_names]

df_subset$Log_Revenue_Total <- log1p(df_subset$Revenue.Total)

# Train the Ridge Regression model using the subset of covariates
ridge_subset_model <- train(Log_Revenue_Total ~ ., 
                            data = df_subset, 
                            method = "glmnet",  
                            trControl = train_control,
                            metric = "RMSE", 
                            preProcess = "scale",  
                            tuneGrid = expand.grid(alpha = 0, lambda = seq(0, 1, length = 100)))

```
```{r}
# Get the coefficients for the best model at the selected lambda
best_lambda <- ridge_model$bestTune$lambda
best_model_coeffs <- coef(ridge_model$finalModel, s = best_lambda)

# Print all coefficients for the best Ridge Regression model
cat("All coefficients for the best Ridge Regression model:\n")
print(best_model_coeffs)

# Get coefficients for the subset model
best_lambda_subset <- ridge_subset_model$bestTune$lambda
best_model_coeffs_subset <- coef(ridge_subset_model$finalModel, s = best_lambda_subset)

# Print all coefficients for the subset model
cat("All coefficients for the Ridge Regression model with a subset of covariates:\n")
print(best_model_coeffs_subset)
```

#### (c)

@@

# Combine p-values into a comparison data frame
comparison <- data.frame(
  Predictor = common_features,
  Training_Pvalues = round(training_pvalues, 8),
  Test_Pvalues = round(test_pvalues, 8)
)

# Sort the comparison data frame in increasing order of Training_Pvalues
comparison <- comparison[order(comparison$Training_Pvalues), ]

@@

# Display the comparison
print("Comparison of p-values from training and test data:")
print(comparison)

## compare coefficients

# Subset coefficients to only include common features
training_coefs <- training_coefs[common_features, , drop = FALSE]
holdout_coefs <- holdout_coefs[common_features, , drop = FALSE]

# Combine coefficients into a comparison data frame
comparison <- data.frame(
  Predictor = common_features,
  Training_Coefs = round(as.numeric(training_coefs), 8),
  Holdout_Coefs = round(as.numeric(holdout_coefs), 8)
)

# Sort the comparison data frame in decreasing order of Training_Coefs
comparison <- comparison[order(-comparison$Training_Coefs), ]

# Display the comparison
print("Comparison of coefficients from training and holdout data:")
print(comparison)

@@

#### (d)

@@

#### (e)

@@

#### GRAVEYARD

```{r 2c Fit to test data Ridge}

# manually refit final ridge model

# Refit the model manually using glmnet
X_train <- model.matrix(Log_Revenue_Total ~ ., data = df_pred)[, -1]
y_train <- df_pred$Log_Revenue_Total

final_ridge_model <- glmnet(X_train, y_train, alpha = 0, lambda = final_lambda_ridge, standardize = TRUE)

# save a copy of the holdout dataset to be transformed/cleaned 
# Create df_test with only the specified columns from holdout
df_test <- holdout[, column_names]

# Extract predictors (X) and response variable (y) for the test set
X_test <- model.matrix(Log_Revenue_Total ~ ., data = df_test)[, -1]
y_test <- df_test$Log_Revenue_Total

# Refit the ridge model on the test set using the final lambda from training
ridge_test <- glmnet(X_test, y_test, alpha = 0, lambda = final_lambda_ridge, standardize = TRUE)

# Extract coefficients from the test set
ridge_test_coefs <- coef(ridge_test)

# Extract coefficients from training and holdout models
training_coefs <- as.matrix(coef(final_ridge_model, s = final_lambda_ridge))
holdout_coefs <- as.matrix(ridge_test_coefs)

# Remove the intercept term (row with name "(Intercept)") from both coefficient matrices
training_coefs <- training_coefs[rownames(training_coefs) != "(Intercept)", , drop = FALSE]
holdout_coefs <- holdout_coefs[rownames(holdout_coefs) != "(Intercept)", , drop = FALSE]

# Ensure feature names match between the training and holdout coefficients
common_features <- intersect(rownames(training_coefs), rownames(holdout_coefs))

# Subset coefficients to only include common features
training_coefs <- training_coefs[common_features, , drop = FALSE]
holdout_coefs <- holdout_coefs[common_features, , drop = FALSE]

# Combine coefficients into a comparison data frame
comparison <- data.frame(
  Predictor = common_features,
  Training_Coefs = round(as.numeric(training_coefs), 8),
  Holdout_Coefs = round(as.numeric(holdout_coefs), 8)
)

# Sort the comparison data frame in decreasing order of Training_Coefs
comparison <- comparison[order(-comparison$Training_Coefs), ]

# Display the comparison
print("Comparison of coefficients from training and holdout data:")
print(comparison)

```{r 2d bootstrap Ridge}

# Function to fit Ridge on bootstrap sample and return coefficients
bootstrap_ridge <- function(data, indices, lambda) {
  # Bootstrap sampling
  boot_sample <- data[indices, ]

  # Create the model matrix from the bootstrap sample
  X_boot <- model.matrix(Log_Revenue_Total ~ ., data = boot_sample)[, -1]
  y_boot <- boot_sample$Log_Revenue_Total

  # Ensure that X_boot has the same column names as the original data
  full_X <- model.matrix(Log_Revenue_Total ~ ., data = data)[, -1]
  full_coefs <- numeric(ncol(full_X))  # Create a vector for all predictors (set to 0 initially)
  names(full_coefs) <- colnames(full_X)  # Name it with the original predictor names

  # Fit Ridge model
  ridge_fit <- glmnet(X_boot, y_boot, alpha = 0, lambda = lambda, standardize = TRUE)

  # Extract coefficients (excluding intercept)
  coefs <- as.vector(coef(ridge_fit)[-1])  # Exclude intercept
  coef_names <- colnames(X_boot)  # Names of predictors in this bootstrap sample

  # Assign the coefficients to the correct positions
  full_coefs[coef_names] <- coefs

  return(full_coefs)
}

# Set up bootstrap
set.seed(123)
n_boot <- 1000  # Number of bootstrap samples
X_train <- model.matrix(Log_Revenue_Total ~ ., data = df_pred)[, -1]
y_train <- df_pred$Log_Revenue_Total

# Perform bootstrap
bootstrap_results <- boot(
  data = df_pred,
  statistic = bootstrap_ridge,
  R = n_boot,
  lambda = final_lambda_ridge
)

# Format as a data frame
cis_df <- data.frame(
  Predictor = colnames(X_train),
  CI_Lower = bootstrap_cis[1, ],
  CI_Upper = bootstrap_cis[2, ]
)

## Compare to the confidence intervals computed in the standard regression output

# Identify numeric predictors
numeric_predictors <- sapply(df_pred[column_names], is.numeric)

# Standardize only numeric predictors
df_pred_standardized <- df_pred
df_pred_standardized[names(numeric_predictors)[numeric_predictors]] <- scale(
  df_pred[names(numeric_predictors)[numeric_predictors]]
)

# # See that non-numeric columns remain unchanged
# print("Standardized dataset:")
# print(head(df_pred_standardized))

# Fit OLS regression with standardized predictors
ols_model_standardized <- lm(Log_Revenue_Total ~ ., data = df_pred_standardized)

# Extract OLS confidence intervals
ols_cis_standardized <- confint(ols_model_standardized, level = 0.95)
ols_cis_df_standardized <- data.frame(
  Predictor = rownames(ols_cis_standardized),
  OLS_CI_Lower_Standardized = ols_cis_standardized[, 1],
  OLS_CI_Upper_Standardized = ols_cis_standardized[, 2]
)

# Merge with Bootstrap confidence intervals for comparison
comparison_ridge_OLS <- merge(
  cis_df,  # Bootstrap confidence intervals from Ridge
  ols_cis_df_standardized,
  by = "Predictor",
  all.x = TRUE
)

# Display the comparison
print("Comparison of Bootstrap and Standardized OLS Confidence Intervals:")
print(comparison_ridge_OLS)

# visualize the difference in Ridge vs. OLS CIs

# Calculate confidence interval sizes for Ridge and OLS
comparison_ridge_OLS$Ridge_CI_Size <- comparison_ridge_OLS$CI_Upper - comparison_ridge_OLS$CI_Lower
comparison_ridge_OLS$OLS_CI_Size <- comparison_ridge_OLS$OLS_CI_Upper_Standardized - comparison_ridge_OLS$OLS_CI_Lower_Standardized

# Reshape the data for plotting
ci_data_ridge <- comparison_ridge_OLS %>%
  select(Predictor, Ridge_CI_Size, OLS_CI_Size) %>%
  pivot_longer(cols = c(Ridge_CI_Size, OLS_CI_Size), 
               names_to = "Method", 
               values_to = "CI_Size")

# Rename methods for better readability
ci_data_ridge$Method <- ifelse(ci_data_ridge$Method == "Ridge_CI_Size", "Ridge", "OLS")

# Create the density plot
ggplot(ci_data_ridge, aes(x = CI_Size, fill = Method)) +
  geom_density(alpha = 0.5) +
  scale_x_log10() +
  labs(
    title = "Density Plot of Confidence Interval Sizes: Ridge vs. OLS",
    x = "Confidence Interval Size",
    y = "Density"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    legend.title = element_blank()
  )

# violin plot
ggplot(ci_data_ridge, aes(x = Method, y = CI_Size)) +
  geom_violin(fill = "lightblue") +
  geom_boxplot(width = 0.1, fill = "white") +
  theme_minimal() +
  labs(title = "Distribution of CI_Size by Group", x = "Group", y = "CI_Size")

```{r 2c Fit to test data Lasso}

# save a copy of the holdout dataset to be transformed/cleaned 
# Create df_test with only the specified columns from holdout
df_test <- holdout[, column_names]

# # Extract predictors (X) and response variable (y) for the test set
# X_test <- model.matrix(Log_Revenue_Total ~ ., data = df_test)[, -1]
# y_test <- df_test$Log_Revenue_Total

# Refit the Lasso model on the test set using the final lambda from training
lasso_test <- glmnet(X_test, y_test, alpha = 1, lambda = final_lambda_lasso, standardize = TRUE)

# Extract coefficients from the test set
lasso_test_coefs <- coef(lasso_test)

# Extract coefficients from training and holdout models
training_coefs <- as.matrix(coef(lasso_model$finalModel, s = final_lambda_lasso))
holdout_coefs <- as.matrix(lasso_test_coefs)

# Remove the intercept term (row with name "(Intercept)") from both coefficient matrices
training_coefs <- training_coefs[rownames(training_coefs) != "(Intercept)", , drop = FALSE]
holdout_coefs <- holdout_coefs[rownames(holdout_coefs) != "(Intercept)", , drop = FALSE]

# Ensure feature names match between the training and holdout coefficients
common_features <- intersect(rownames(training_coefs), rownames(holdout_coefs))

# Subset coefficients to only include common features
training_coefs <- training_coefs[common_features, , drop = FALSE]
holdout_coefs <- holdout_coefs[common_features, , drop = FALSE]

# Combine coefficients into a comparison data frame
comparison <- data.frame(
  Predictor = common_features,
  Training_Coefs = round(as.numeric(training_coefs), 8),
  Holdout_Coefs = round(as.numeric(holdout_coefs), 8)
)

# Sort the comparison data frame in decreasing order of Training_Coefs
comparison <- comparison[order(-comparison$Training_Coefs), ]

# Display the comparison
print("Comparison of coefficients from training and holdout data:")
print(comparison)

```{r 2d bootstrap Lasso}

# Function to fit Lasso on bootstrap sample and return coefficients
bootstrap_lasso <- function(data, indices, lambda) {
  # Bootstrap sampling
  boot_sample <- data[indices, ]

  # Create the model matrix from the bootstrap sample
  X_boot <- model.matrix(Log_Revenue_Total ~ ., data = boot_sample)[, -1]
  y_boot <- boot_sample$Log_Revenue_Total

  # Ensure that X_boot has the same column names as the original data
  full_X <- model.matrix(Log_Revenue_Total ~ ., data = data)[, -1]
  full_coefs <- numeric(ncol(full_X))  # Create a vector for all predictors (set to 0 initially)
  names(full_coefs) <- colnames(full_X)  # Name it with the original predictor names

  # Fit Lasso model
  lasso_fit <- glmnet(X_boot, y_boot, alpha = 1, lambda = lambda, standardize = TRUE)

  # Extract coefficients (excluding intercept)
  coefs <- as.vector(coef(lasso_fit)[-1])  # Exclude intercept
  coef_names <- colnames(X_boot)  # Names of predictors in this bootstrap sample

  # Assign the coefficients to the correct positions
  full_coefs[coef_names] <- coefs

  return(full_coefs)
}

# Set up bootstrap
set.seed(123)
n_boot <- 1000  # Number of bootstrap samples
X_train <- model.matrix(Log_Revenue_Total ~ ., data = df_pred)[, -1]
y_train <- df_pred$Log_Revenue_Total

# Perform bootstrap
bootstrap_results <- boot(
  data = df_pred,
  statistic = bootstrap_lasso,
  R = n_boot,
  lambda = final_lambda_lasso
)

# Format as a data frame
cis_df <- data.frame(
  Predictor = colnames(X_train),
  CI_Lower = bootstrap_cis[1, ],
  CI_Upper = bootstrap_cis[2, ]
)

## Compare to the confidence intervals computed in the standard regression output

# Identify numeric predictors
numeric_predictors <- sapply(df_pred[column_names], is.numeric)

# Standardize only numeric predictors
df_pred_standardized <- df_pred
df_pred_standardized[names(numeric_predictors)[numeric_predictors]] <- scale(
  df_pred[names(numeric_predictors)[numeric_predictors]]
)

# # See that non-numeric columns remain unchanged
# print("Standardized dataset:")
# print(head(df_pred_standardized))

# Fit OLS regression with standardized predictors
ols_model_standardized <- lm(Log_Revenue_Total ~ ., data = df_pred_standardized)

# Extract OLS confidence intervals
ols_cis_standardized <- confint(ols_model_standardized, level = 0.95)
ols_cis_df_standardized <- data.frame(
  Predictor = rownames(ols_cis_standardized),
  OLS_CI_Lower_Standardized = ols_cis_standardized[, 1],
  OLS_CI_Upper_Standardized = ols_cis_standardized[, 2]
)

# Merge with Bootstrap confidence intervals for comparison
comparison_lasso_OLS <- merge(
  cis_df,  # Bootstrap confidence intervals from Lasso
  ols_cis_df_standardized,
  by = "Predictor",
  all.x = TRUE
)

# Display the comparison
print("Comparison of Bootstrap and Standardized OLS Confidence Intervals:")
print(comparison_lasso_OLS)

# visualize the difference in Lasso vs. OLS CIs

# Calculate confidence interval sizes for Lasso and OLS
comparison_lasso_OLS$Lasso_CI_Size <- comparison_lasso_OLS$CI_Upper - comparison_lasso_OLS$CI_Lower
comparison_lasso_OLS$OLS_CI_Size <- comparison_lasso_OLS$OLS_CI_Upper_Standardized - comparison_lasso_OLS$OLS_CI_Lower_Standardized

# Reshape the data for plotting
ci_data_lasso <- comparison_lasso_OLS %>%
  select(Predictor, Lasso_CI_Size, OLS_CI_Size) %>%
  pivot_longer(cols = c(Lasso_CI_Size, OLS_CI_Size), 
               names_to = "Method", 
               values_to = "CI_Size")

# Rename methods for better readability
ci_data_lasso$Method <- ifelse(ci_data_lasso$Method == "Lasso_CI_Size", "Lasso", "OLS")

# Create the density plot
ggplot(ci_data_lasso, aes(x = CI_Size, fill = Method)) +
  geom_density(alpha = 0.5) +
  scale_x_log10() +
  labs(
    title = "Density Plot of Confidence Interval Sizes: Lasso vs. OLS",
    x = "Confidence Interval Size",
    y = "Density"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    legend.title = element_blank()
  )

# violin plot
ggplot(ci_data_lasso, aes(x = Method, y = CI_Size)) +
  geom_violin(fill = "lightblue") +
  geom_boxplot(width = 0.1, fill = "white") +
  theme_minimal() +
  labs(title = "Distribution of CI_Size by Group", x = "Group", y = "CI_Size")

>>>>

### Statistically Significant Coefficients
```{r 2a}
# Display summary of the OLS model
summary_ols <- summary(ols_train)

# Extract coefficients table
coefficients_table <- summary_ols$coefficients

# Identify statistically significant coefficients
significant_coefficients <- coefficients_table[coefficients_table[, 4] < 0.05, ]

# Print significant coefficients
print(significant_coefficients)

@@

### Subset Model
```{r}
# Select a subset of covariates
# Subset the dataframe
subset_column_names <- c("Demand.Summer.Peak", "Demand.Winter.Peak", "Sources.Generation", 
                         "Utility.Region", "Max.Revenue.Sector", "Revenue.Total", "Utility.Type")
df_subset <- df[, subset_column_names]
df_subset$Log_Revenue_Total <- log1p(df_subset$Revenue.Total)

# Create the formula for the subset model
subset_formula <- as.formula("Log_Revenue_Total ~ Demand.Summer.Peak + Demand.Winter.Peak + 
                              Sources.Generation + Utility.Region + Max.Revenue.Sector + Utility.Type")

# Fit the subset OLS model
subset_ols <- lm(subset_formula, data = df_subset)

# Print the summary of the subset model
summary(subset_ols)

```{r}
# Extract significant coefficients from the subset model
subset_summary <- summary(subset_ols)
subset_coefficients <- subset_summary$coefficients
significant_subset_coefficients <- subset_coefficients[subset_coefficients[, 4] < 0.05, ]

# Print significant coefficients from the subset model
print(significant_subset_coefficients)
```

### OLS with test data

@@

# Combine p-values into a comparison data frame
comparison_pvalues <- data.frame(
  Predictor = common_features,
  Training_Pvalues = round(training_pvalues, 8),
  Test_Pvalues = round(test_pvalues, 8)
)

# Sort the comparison data frame in increasing order of Training_Pvalues
comparison_pvalues <- comparison_pvalues[order(comparison_pvalues$Training_Pvalues), ]

@@

# Display the comparison of p-values
print("Comparison of p-values from training and test data:")
print(comparison_pvalues)

# Extract coefficients from the training and test models
training_coefs <- coef(ols_train)
holdout_coefs <- coef(ols_test)

# Ensure feature names match between training and holdout coefficients
common_features <- intersect(names(training_coefs), names(holdout_coefs))

# Subset coefficients to only include common features
training_coefs <- training_coefs[common_features]
holdout_coefs <- holdout_coefs[common_features]

# Combine coefficients into a comparison data frame
comparison_coefs <- data.frame(
  Predictor = common_features,
  Training_Coefs = round(as.numeric(training_coefs), 8),
  Holdout_Coefs = round(as.numeric(holdout_coefs), 8)
)

# Sort the comparison data frame in decreasing order of Training_Coefs
comparison_coefs <- comparison_coefs[order(-comparison_coefs$Training_Coefs), ]

# Display the comparison of coefficients
print("Comparison of coefficients from training and holdout data:")
print(comparison_coefs)

@@

### Bootstrap

@@

### Multiple Hypothesis Testing

@@
